---
title: "Coursework for Survival Models"
author: "Wanting Xu (CID:01508313)"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
  word_document: default
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(survival);library(ggplot2);library(ranger);library(ggfortify);
library(flexsurv);library(pch);library(dplyr);library(survminer)
```

## 1.(a)
 
#### Step 1. Generating a data set with 40% censored observations
```{r}
S0 <- function(t){
  pgamma(t, shape = 2, rate = 1.8, lower.tail = FALSE)}

set.seed(109)
n <- 200
T.pop <- rgamma(n = n, shape = 2, rate = 1.8) # a data set required to be simulated
C <- rweibull(n = n,shape = 1,scale = 2) # censoring times
T.obs <- pmin(T.pop, C)
D <- ifelse(T.pop <= C, 1, 0)
table(D) # Check the percentage of censoring
```

Here, we have 79 '0's in D, $\frac{79}{200}=39.5\%\approx 40\%.$

```{r}
S.comp <- Surv(time = T.pop, event = rep(1, n))
S.cens <- Surv(time = T.obs, event = D)
KM <- survfit(S.cens ~ 1)
eS <- survfit(S.comp ~ 1)
survfit(S.cens ~ 1)
```

In dataset A with 200 samples in total, 79 samples were censored. Among 121 data, median survival time = 0.873.

The **95% confidence interval** for 121
uncensored individuals is (0.776, 1.05).
  
  
#### Step 2. Plot the Kaplan-Meirer estimate with confidence intervals.
```{r}
layout(rbind(1:2))
curve(S0, 0, 5, col=1, lwd=2, ylab="S(t)", xlab="t"); lines(eS, col=3, lwd=2, conf.int=TRUE)
title(main="Empirical survival curve", sub="(uncensored data)")
curve(S0, 0, 5, col=1, lwd=2, ylab="S(t)", xlab="t"); lines(KM, col=4, lwd=2, conf.int=TRUE)
title(main="Kaplan-Meier survival curve", sub="(right-censored data)")
```

We observed large jumps for large t.
 
## 1b
```{r}
Cb <- T.pop+3 # Generating a time-dependent 'C'
Db <- ifelse(T.pop <= Cb, 1, 0)
S.compb <- Surv(time = T.pop, event = rep(1, n))
S.censb <- Surv(time = pmin(T.pop, Cb), event = D)
layout(rbind(1:2))
curve(S0, 0, 5, col=1, lwd=2, ylab="S(t)", xlab="t")
lines(survfit(S.compb ~ 1), col=3, lwd=2, conf.int=TRUE); title(main="Empirical survival curve")
curve(S0, 0, 5, col=1, lwd=2, ylab="S(t)", xlab="t")
lines(survfit(S.censb ~ 1), col=4, lwd=2, conf.int=TRUE); title(main="Kaplan-Meier survival curve")
```
 
## 1c

Modelling the distribution of step function
```{r}
gendat <- function(){n <- 200
  T.pop <- rgamma(n = n, shape = 2, rate = 1.8)
  C <- rweibull(n = n,shape = 1,scale = 2)
  T.obs <- pmin(T.pop, C)
  D <- ifelse(T.pop <= C, 1, 0)
  list(X=T.obs, Delta = D)}
data <-gendat() 

model <- list(
  validtheta= function(theta) theta>0, # only postive rate
  h=function(x,theta) rep(theta,length(x)), # hazard func
  H=function(x,theta) theta*x # integrate from constant to x
  )
```
  
### Maximum Likelihood Estimation in 1-dim
Check the log likelihood function
```{r}
l <- function(theta,data) {
  if (!model$validtheta(theta)) return(-Inf)   # checking theta is valid, if not, return -inf
  sum(log(model$h(data$X[data$Delta==1],theta)))-sum(model$H(data$X,theta))} #picking out all uncensored
theta <- seq(0,5,by=0.01)
plot(theta,sapply(theta,function(y) l(y,data)),ylab=expression("log-likelihood"*(theta)),xlab=expression(theta),type="l")
```

By above graph, we could predict the maximum value appears in $(0.5,1)$.
 
Then we work out the parameter in 1d optimisation.
```{r}
o1 <- optim(c(1),fn=function(theta) -l(theta,data), method="Brent",lower=1e-4,upper=1e6, hessian=TRUE)
o1$par
```
-  Par value looks close to the maximum point of the curve.
-  When tried a larger sample size = 2000, par = 0.773 which is closer to true maximum.
-  Convergence = 0: we could trust the result. 
```{r}
sqrt(solve(o1$hessian)) 
```

```{r}
# check whether the result is reasonable. try 1000 times
thetas <- replicate(1000,{ data <- gendat()
  optim(c(1),fn=function(theta)-l(theta,data),method="Brent",lower=1e-4,upper=1e6,hessian=TRUE)$par})
sd(thetas)
```
These 2 sd are similar. We can use sd and par to calculate confidence intervals.
### Calculating 95% confidence intervals
```{r}
CI <- o1$par + diag(solve(o1$hessian))*qnorm(c(0.025,0.975))
CIs <- replicate(10000,{ data <- gendat()
  o2 <- optim(c(1),fn=function(theta)-l(theta,data),method="Brent",lower=1e-4,upper=1e6,hessian=TRUE)
  o2$par + sqrt(as.numeric(solve(o2$hessian)))*qnorm(c(0.025,0.975))
} ) # check
mean(0.74>=CIs[1,] & 0.74<=CIs[2,])
```
95% of these confidence intervals contain the true value. When we try a large n, this ratio is closer to 95%.
  
### Modelling with the piecewise step function
 
#### Hazard rate and integrated hazard rate
I divide dataset into 5 equal intervals.
```{r}
# Piecewise step func with constant hazard rate in each time interval
model <- list( validtheta= function (theta) all(theta>0),
  h=function(x,theta) 
    ifelse(x<=KM$time[40], theta[1], ifelse(x<=KM$time[80],theta[2],
                                                ifelse(x<=KM$time[120],theta[3], ifelse(x<=KM$time[160],theta[4], theta[5])))),
  # Integrate the hazard rate function
  H=function(x,theta) 
    ifelse(x<=KM$time[40], x*theta[1],
           ifelse(x<=KM$time[80],KM$time[40]*theta[1]+(x-KM$time[40])*theta[2],
                  ifelse(x<=KM$time[120], KM$time[40]*theta[1]+(KM$time[80]-KM$time[40])*theta[2]+(x-KM$time[80])*theta[3],
                         ifelse(x<=KM$time[160], KM$time[40]*theta[1]+(KM$time[80]-KM$time[40])*theta[2]+(KM$time[120]-KM$time[80])*theta[3]+(x-KM$time[120])*theta[4],
                                KM$time[40]*theta[1]+(KM$time[80]-KM$time[40])*theta[2]+(KM$time[120]-KM$time[80])*theta[3]+(KM$time[160]-KM$time[120])*theta[4]+(x-KM$time[160])*theta[5])))))
```
  
#### Optimising the theta in piecewise function
```{r}
data <- gendat()
o5 <- optim(c(1,1,1,1,1),fn=function(theta) -l(theta,data),hessian=TRUE)
sd <- sqrt(diag(solve(o5$hessian)))
```
 
#### Confidence intervals in piecewise function
```{r}
CIs = list(o5$par[1]+sd[1]*qnorm(c(0.025,0.975)),o5$par[2]+sd[2]*qnorm(c(0.025,0.975)),o5$par[3]+sd[3]*qnorm(c(0.025,0.975)),o5$par[4]+sd[4]*qnorm(c(0.025,0.975)),o5$par[5]+sd[5]*qnorm(c(0.025,0.975)))
CI1 <- c(CIs[[1]][1],CIs[[2]][1],CIs[[3]][1],CIs[[4]][1],CIs[[5]][1])
CI2 <- c(CIs[[1]][2],CIs[[2]][2],CIs[[3]][2],CIs[[4]][2],CIs[[5]][2])
```
  
#### Plotting graph
```{r}
htrue <- function(t){ hgamma(t,shape = 2, rate = 1.8)} # True hazard

x1 <- c(KM$time[40],KM$time[80],KM$time[120],KM$time[160])
f1 <- stepfun(x1,o5$par,f=0)
f_lower <- stepfun(x1,CI1,f=0)
f_upper <- stepfun(x1,CI2,f=0)

x<-seq(from = 0, to = 1.5, length.out = 1000)
plot(x,htrue(x),type='l',lty=1,lwd=2,col=4,xlim=c(0,1.5),ylim=c(0,2),xlab="time",ylab='hazard value')
curve(f1,lty=3,lwd=2,col=1,add=TRUE)
curve(f_upper,lty=2,lwd=2,col=3,add=TRUE)
curve(f_lower,lty=2,lwd=2,col=2,add=TRUE)
legend("bottomright",c("true hazard rate","estimated hazard rate","asymptotic 95% upper CI","asymptotic 95% lower CI"),col=c(4,1,3,2),text.font=1,lty=1:2,cex=0.8)
title(main="Estimated Piecewise Hazard Rate")
```

## 2

I use **vereran** dataset from the 'survival' package in R, which is the randomised trial of two treatment regimens for lung cancer.
<https://www.rdocumentation.org/packages/survival/versions/3.2-13/topics/veteran>

-   trt: '1' = standard, '2' = test
-   time: days of survival 
-   diagtime: months from diagnosis 
-   status: censoring status: '= 1' an observed
death, '=0' censored data 
-   karno: performnance status 
-   prior: '0' no prior therapy, '10' has prior therapy

```{r}
head(veteran)  #view the data
veteran$trt <- factor(veteran$trt, labels = c("standard", "test"))
veteran$prior <- factor(veteran$prior, labels = c("no", "yes"))
veteran$status <- as.factor(veteran$status)
# summary(veteran)
```

The above summary shows, (because of page limit, it doesn't show up)

-   The **range of age** is $[34,81]$.
-   The diagnosis time is obviously **right skewed**.
-   $\frac{9}{137}=6.6\%$ observations are censored.
    $\frac{128}{137}=93.4\%$ of patients died before the end of the study.
-   $\frac{40}{137}=29.2\%$ patients had a prior therapy.
 
### 1) Survival curves without covariates by Kaplan-Meier analysis
```{r}
rm(veteran)
KMvet <- with(veteran, Surv(time, status))
vet_fit <- survfit(Surv(time, status) ~ 1, data=veteran)
ggsurvplot(fit = vet_fit, data = veteran, title = "Overall Survival", font.title = c(15, "bold"), 
           ggtheme = theme_minimal() + theme(plot.title = element_text(hjust = 0.5, face = "bold")),
           xlab="Days", ylab = "Survival Probability", font.x=c(12,"bold"), font.y=c(15,"bold"),
           font.xtickslab=c(8,"plain"), font.ytickslab=c(8,"plain"),
           censor.shape="|", censor.size = 4, conf.int = TRUE, conf.int.fill = "purple",
           surv.median.line = "hv", legend.title = "All Patients", legend.labs = "All Patients",
           risk.table = TRUE, risk.table.height = 0.25, risk.table.fontsize = 4.5)
```

### 2) Multivariate model by Cox
```{r}
cox_mult <- coxph(Surv(time, status) ~ ., data = veteran)
summary(cox_mult)
```

-   The 3 p-values are similar. The model is **statistically significant**
    (p-value \< 0.05).
-   **Cell type and karno probably are the most significant variables.**

```{r}
ggsurvplot(survfit(cox_mult, data = veteran), ggtheme = theme_minimal())
```
 
### 3) Analysis with covariates: treatment
 
This set has right-skewed property as mentioned before.
 
#### Kaplan-Meier analysis for treatment
```{r}
KMvet_trt_fit <- survfit(Surv(time, status) ~ trt, data=veteran)
head(KMvet_trt_fit)

ggsurvplot(KMvet_trt_fit, data = veteran, pval = TRUE, conf.int = TRUE, break.time.by = 150,
  xlab = "Time in days", ggtheme = theme_light(), risk.table = "abs_pct", risk.table.y.text.col = T,
  risk.table.y.text = FALSE, ncensor.plot = TRUE, surv.median.line = "hv",  # median survival
  legend.title = "trt types")
```

-   p = 0.93: extremly high, there is insufficent evidence to reject $H_0$. So the treatment type doesn't influecne the survival time.
-   For standard treatment, median survival time = 103 days, while for test treatment, median survival time = 52.5 days (much lower).
-   However, because of a high p-value, although medians have a large difference, there is no strong evidence to claim curves are different.
 
#### Fitting a Cox Model
```{r}
# summary(coxph(Surv(time,status)~trt,data=veteran))
```
Because of page limit, the summary doesn't show up

-   $Coef = 0.01774$: test patients have higher risk of death than standard patients.
-   $exp(coef) = 1.018$: having test treatment increases the hazard by a
    factor of 1.02.
-   $[0.714,1.45]$: 95% confidence intervals of hazard ratio.

```{r}
survreg(Surv(time,status)~trt,data=veteran)
```
  
#### Log-Rank Analysis
```{r}
survdiff(Surv(time = veteran$time, event = veteran$status) ~ veteran$trt, data = veteran)
```

The observed and expected values in both groups of patients are very
similar. And $p = 0.9 \approx 1.0$
  
### 4) Analysis with covariates: cell type

```{r}
vet_surv <- Surv(time = veteran$time, event = veteran$status)
vet_celltype <- survfit(vet_surv ~ veteran$celltype, data = veteran)
summary(vet_celltype, times = c(1, 50, 100*(1:6))) # View data at various periods (600-900 have same value)
```

```{r}
ggsurvplot(fit = vet_celltype, data = veteran,
           title = "Overall Survival analysed by celltypes", font.title = c(15, "bold"),
           ggtheme = theme_minimal() + theme(plot.title = element_text(hjust = 0.5, face = "bold")),
           censor = TRUE, censor.shape="|", censor.size = 5, conf.int = TRUE, xlab="Days",
           surv.median.line = "hv",  ylab = "Survival Probability", font.x=c(12,"bold"), font.y=c(12,"bold"), 
           font.xtickslab=c(10,"plain"), font.ytickslab=c(10,"plain"), legend = "none", legend.title = "All Patients", 
           legend.labs = c("Squamous","Small Cell","Adeno","Large"), surv.plot.height = 0.85,
           risk.table = TRUE, risk.table.height = 0.25, risk.table.fontsize = 3)
```
 
#### Log-Rank Analysis
```{r}
survdiff(Surv(time = veteran$time, event = veteran$status) ~ veteran$celltype, data = veteran)
```

Patients with Small Cell and Adeno are observed more events than
expected values, therefore they perform worse than expectation.
  
#### Fitting a Cox Model
```{r}
coxph(Surv(time = veteran$time, event = veteran$status)  ~ veteran$celltype, data = veteran)
```

-   The types of small cell and adeno are significant
-   By the above graph, the area of Large Cell has a better performance,
    it may good for patients.
  
**Aim**: Test whether the treatment is effective in Small Cell Lung
Cancer. 

**Method**: export the subset, then do a log-rank test or Cox.
```{r}
# log-rank test (include=FALSE)
vet_small <-veteran %>% filter(celltype=="smallcell")
survdiff(Surv(time = vet_small$time, event = vet_small$status) ~ vet_small$trt, data = vet_small)
```

```{r}
# CoxPH analysis
coxph(Surv(time = vet_small$time, event = vet_small$status) ~ vet_small$trt, data = vet_small)
```

We could apply same method to test the performance of the treatment for
other cell types.
 
### 5) Analysis with covariates: age

The median age in the dataset is 62 yrs old. We classified people with $age<62$ *(LT62)* as young people, and others are the olds *(GT62)*.
```{r}
vet <- mutate(veteran, AGE = ifelse((age < 62), "age<62", "age>62"), AGE = factor(AGE), trt = factor(trt,labels=c("standard","test")), prior = factor(prior,labels=c("N0","Yes")))
KM_AGE_fit <- survfit(Surv(time, status) ~ AGE, data=vet)
autoplot(KM_AGE_fit)
```

Clearly, the two curves almost overlap in the first 100 days. **Younger patients are more likely to survive more than a year.**
 
### Visualising Hazard Ratios

```{r message=FALSE, warning=FALSE}
model.vet <- coxph(Surv(time, status) ~ karno + diagtime + age, data = veteran )
ggforest(model.vet,cpositions = c(0.05, 0.2, 0.4),fontsize = 0.6)
```
